%
% file: dedication.tex
% description: Contains the text for thesis dedication
%

\chapter*{Dedication and acknowledgements}
\begin{SingleSpace}
\initial{I}n memory of Peter Feighan, who is missed every day.

\hfill \break
\hfill \break
\hfill \break

This thesis would not be possible with the help and support of a great number of
people, to whom I am very thankful.

\hfill \break

Foremost are my supervisors, Prof. Fred Manby and Prof. Andrew Orr-Ewing, whose
helpful advice on all elements of my project, from scientific challenges to practical
skills, has been invaluable and without whom this project would not exist or have 
been finished.

\hfill \break

Equally I am very grateful to Dr. Susannah Bourne-Worster, who has provided so
much time and effort in helping to realizing the full potential of these projects
and others in the course of my PhD. I am deeply grateful for her support with all 
levels of this work.

\hfill \break

I am indebted to the members of the Manby research group who taught me the technical
skills necessary to carry out this work, especially Dr. Peter Bygrave, Dr. James
Womack and Dr. Thomas Dresselhaus. I have greatly enjoyed the supportive and friendly
atmosphere of this group, and the University of Bristol's Centre for Computational 
Chemistry in general - this environment has provided many stimulating conversations 
and I have found amiable encouragement through tough periods, especially with the 
support from Harry Stroud and Joe Crossley-Lewis.

\hfill \break

Finally I would like to thank my family and friends, whose presence has been inspirational
and set the foundations that led me to this work.

\end{SingleSpace}
\clearpage